---
title: "Data Wrangling"
author: "Abhishek Dendukuri"
date: "7/7/2018"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(car)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

First we set the working directory and read in the information from our properties file into a dataframe. This spreadsheet proves to be massive in size so the data frame generation tends to take a few minutes before everything is properly setup
``` {r}
setwd("/Users/JARVIS/RStudio/Capstone")
df <- read.csv("properties_2017.csv")
properties <- tbl_df(df)
```

Now that everything is set up we can see that the table has generated and there are more than two million rows worth of values
``` {r}
str(properties)
```

## Missing Percentage Plot

In order to understand which information to remove, it is best to display what factors are missing the highest amount of values. From here we can create a cutoff that would represent our data in a tidier fashion

```{r fig1, fig.height=10}
missing_values <- properties %>% summarize_all(funs(sum(is.na(.)/n())))
missing_values <- gather(missing_values, key="feature", value="missing_pct")
missing_values %>% 
  ggplot(aes(x=reorder(feature,-missing_pct),y=missing_pct)) +
  geom_bar(stat="identity",fill="red")+
  coord_flip()+theme_bw()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
