---
title: "Data Story"
author: "Abhishek Dendukuri"
date: "7/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Housing prices have long been dictated and settled by real estate agents as their background knowledge about the property reliably led rto sales. But as techniiology evolves and as new data is foun, it was only time before companies like Zillow attacked the online housing market to track property prices for prospective homewoners.

The Zestimate model that Zillow devloped computs the estimated market value of a home using a formula that only analysts at the company know. Using certain factors and environmental variables, it generates a comparative market analysis - or CMA - that can determina valuation similar to one given by a real estate agent.

```{r}

```

## Deep Dive into the Dataset

What I would like to understand is how well some of the individual factors correlate to the Zestimate as a whole. One of the benefits of having a dataset that Zillow provides is the cinredible breadth of factors as well as the vast number of properties that the Zestmiate analyzes. It is important to have a number of data points in order to create a much more accurate algorithm or formula that can utilize all this information to create a better model. Further analysis reveals that the entire dataset provided only supplies housing data for properties in the Los Angeles area. This is really helpful because we can visualize how the real estate market is for only one small area of the map - anything larger would cause the number of properties to explode, making it harder to organize the data.

However the dataset is not without its drawbacks. Namely, the breadth of factors comes into question to a point where some of the factors seem unnecessary to include because they are missing several data points. This will be explored to a greater extent in the data wrangling section. As mentioned earlier, the data is only available for the Los Angeles area, and while it is helpful that all the data is localized in one region, this can be seen as a drawback because as far as our analysis goes, we will not be able to compare how well the formula applies between states.